---
title: "Final task"
author: "Victoria"
date: "2024-04-18"
output: html_document
---

# Libraries

```{r}
library(impute)
library(limma)
library(ggplot2)
library(vegan)
library(dplyr)
library(tidyverse)
library(readxl)
library(ape)
library(clusterProfiler)
library(dendextend)
library(RColorBrewer)
library(Biobase)
library(gplots)
library(ggrepel)
library(gprofiler2)
library(dendextend)
library(NMF)
library(DT)
```

# Data

```{r}
data <- read.csv("peaks_data.csv")
```

## data preprocessing and normalization

# remove missing values
```{r}
sum(is.na(data))
data <- data[which(rowMeans(!is.na(data)) > 0.5), ]
data <- data[which(!duplicated(data$Gene_id)),]
rownames(data) <- data$Gene_id
sum(is.na(data))
data <- data[, -c(1:3)]
```
# imputation of missed data with KNN
```{r}
data_m <- as.matrix(data)
data_m <- t(data_m)
knn_d <- impute.knn(data_m, k = 5)
data_m <- knn_d$data
data_m <- as.data.frame(data_m)
data_m <- t(as.matrix(data_m))
head(data_m, 3)
```

# normalizing
```{r}
boxplot(data_m, col = c(rep("blue", 14), rep("red", 19)), main = "data")
legend("bottomright", legend = c("BT", "CJK"), fill = c("blue", "red"))
```

```{r}
data_norm <- log2(data_m + 1)
data_norm <- normalizeQuantiles(as.matrix(data_norm))
```

```{r}
boxplot(data_norm, col = c(rep("blue", 14), rep("red", 19)), main = "Normalized data")
legend("bottomright", legend = c("BT", "CJK"), fill = c("blue", "red"))
```

## Evaluate the presence of a batch-effect using PCA

```{r}
d_pca <- t(data_norm)
rda_pca <- rda(d_pca, scale = TRUE)

rownames(d_pca) <- colnames(data_m)
```

```{r}
d_scores <- data.frame(
    d_pca,
    scores(rda_pca, display = "sites", choices = c(1, 2, 3), scaling = "sites")
)

p_scores <- ggplot(d_scores, aes(x = PC1, y = PC2)) +
    geom_point(aes(color = rownames(d_pca)), alpha = 0.5) +
    coord_equal(xlim = c(-3, 3), ylim = c(-3, 3)) +
    ggtitle(label = "Ordination") +
    theme_bw()
p_scores
```
# MA-plot
```{r}
maplot <- function(X1, X2, pch = 21, main = "MA-plot", xlab = "Average log-expression", ylab = "Expression log-ratio", lpars = list(col = "blue", lwd = 2), ...) {
  X <- (rowMeans(X2) + rowMeans(X1)) / 2
  Y <- rowMeans(X2) - rowMeans(X1)
  scatter.smooth(x = X, y = Y, main = main, pch = pch, xlab = xlab, ylab = ylab, lpars = lpars, ...)
  abline(h = c(-1, 0, 1), lty = c(2, 1, 2))
}
```

```{r}
maplot(data_norm[, 1:14], data_norm[, 14:32])
```
## Identify differentially expressed proteins

```{r}
expr_data <- as.matrix(data_norm)

experiment <- as.factor(sapply(strsplit(colnames(data_m), "_"), `[`, 1))
pheno_data <- data.frame(experiment)
rownames(pheno_data) <- colnames(data_m)
pheno_metadata <- data.frame(
    labelDescription = c("Experimental condition"),
    row.names = c("Condition")
)
pheno_data <- new("AnnotatedDataFrame",
    data = pheno_data,
    varMetadata = pheno_metadata
)

feature_data <- data.frame(Prot = rownames(expr_data))
rownames(feature_data) <- rownames(expr_data)
feature_metadata <- data.frame(
    labelDescription = c("Protain name"),
    row.names = c("Protain")
)
f_data <- new("AnnotatedDataFrame",
    data = feature_data,
    varMetadata = feature_metadata
)

exp_set <-
    ExpressionSet(
        assayData = expr_data,
        phenoData = pheno_data,
        featureData = f_data
    )
```

```{r}
X <- model.matrix(~experiment, pData(exp_set))
fit <- lmFit(exp_set, design = X, method = "robust", maxit = 1000)
efit <- eBayes(fit)
```
```{r}
MA_limma <- function(efit, coef, n = 10, signif = TRUE, fdr = 0.05, lfc = 0, text = TRUE, cex.text = 0.8, col.text = "grey20", main = "MA-plot", xlab = "Average log-expression", ylab = "Expression log-ratio", pch = 19, pch.signif = 21, col = "darkgreen", alpha = 0.3, cex = 0.3, ...){
  
  R <- efit$coefficients[, coef]
  I <- efit$Amean
  
  col_btransp <- adjustcolor(col, alpha.f = alpha)
  
  plot(I, R, cex = cex, main = main, pch = pch, xlab = xlab, ylab = ylab, col = col_btransp, ...)
  abline(h = 0)
  
  if(signif){
    sign <- p.adjust(efit$p.value[, coef], method = "BH") <= fdr
    large <- abs(efit$coefficients[, coef]) >= lfc
    points(I[sign & large], R[sign & large], cex = cex*2, col = "orange2", pch = pch.signif)
  }
  
  if(text){
    ord <- order(efit$lods[, coef], decreasing = TRUE)
    top_n <- ord[1:n]
    text(I[top_n], R[top_n], labels = efit$genes[top_n, ], pos = 4, cex = cex.text, col = col.text)
  }
}


MA_limma(efit, coef = 2, n = 30)
```
```{r}
my_list <- topTable(efit, coef = 2, n = 100)
dif_exp_set <- exp_set[fData(exp_set)$Prot %in% my_list$Prot, ]
```

```{r}
dat <- as.matrix(exprs(dif_exp_set))
pal_blue_red <- colorpanel(75, low = "steelblue", mid = "black", high = "red")
heatmap.2(dat, col = pal_blue_red, scale = "row", key = TRUE, symkey = FALSE, density.info = "none", trace = "none", cexRow = 0.9, cexCol = 1, margins = c(4, 3), keysize = 0.8, key.par = list(mar = c(3, 0.1, 3, 0.1)))

```

```{r}
topTable(efit, coef = 2)
numGenes <- nrow(exprs(exp_set))
full_list <- topTable(efit, number = numGenes)
full_list <- full_list[full_list$adj.P.Val <= 0.05, ]
#write.csv(full_list, "DE_full.csv")
```

```{r}
MA_limma(efit, coef = 2, n = 80, text = F, lfc = 1)
```

```{r}
my_list <- full_list
dif_exp_set <- exp_set[fData(exp_set)$Prot %in% my_list$Prot, ]
```
# vulcano plot
```{r }
volcano_list <- full_list

volcano1 <- ggplot(data = volcano_list, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()

volcano2 <- volcano1 + geom_vline(xintercept=c(-0.6, 0.6), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red")

volcano_list$diffexpressed <- "NO"       
```
```{r }
volcano_list$diffexpressed[volcano_list$logFC > 1 & volcano_list$adj.P.Val < 0.05] <- "UP"
volcano_list$diffexpressed[volcano_list$logFC < -1 & volcano_list$adj.P.Val< 0.05] <- "DOWN"

volcanodif1 <- ggplot(data = volcano_list, aes(x = logFC, y = -log10(adj.P.Val), col = diffexpressed)) +
    geom_point() +
    theme_minimal() +
    geom_text_repel(data = head(volcano_list, 20), aes(label = head(volcano_list, 20)$Prot))
volcanodif1
```
```{r}
diff_prot <- rownames(full_list)
diff_expressed <- as.data.frame(data_norm)[diff_prot, ]
t_diff_expressed <- t(diff_expressed)
# rownames(t_diff_expressed) <-  as.factor(gsub("_[^_]*", replacement = "", rownames(t_diff_expressed)))
# rownames(t_diff_expressed) <- make.unique(as.character(pheno$df_experiment))

d <- dist(x = t_diff_expressed, method = "canberra")

mouse_hc_avg <- hclust(d, method = "average")
mouse_ph_avg <- as.phylo(mouse_hc_avg)
mouse_den_avg <- as.dendrogram(mouse_hc_avg)

get_colours <- function(dend, n_chars, palette = "Dark2") { 
    labs <- get_leaves_attr(dend, "label")
    group <- substr(labs, start = 0, stop = n_chars)
    group <- factor(group)
    cols <- brewer.pal(length(levels(group)), name = palette)[group]
    return(cols)
}

cols <- get_colours(dend = mouse_den_avg, n_chars = 6)
den_avg_c <- color_labels(dend = mouse_den_avg, col = cols)
plot(den_avg_c, horiz = TRUE)
```

# enrichment analysis
```{r}
# keep only the significant proteins results
sig <- subset(full_list, adj.P.Val < 0.05)
# get the significant up-regulated proteins
up <- subset(sig, logFC > 0)
# get the significant down-regulated proteins
down <- subset(sig, logFC < 0)
# needed to convert to enrichResult object
up_names <- gconvert(row.names(up))
down_names <- gconvert(row.names(down))
```
```{r}
multi_gp_up_reg <- gost(list("up-regulated" = up_names$name), multi_query = FALSE, evcodes =TRUE)
gp_mod_up = multi_gp_up_reg$result[, c("query", "source", "term_id","term_name", "p_value", "query_size", "intersection_size", "term_size", "effective_domain_size", "intersection")]
gp_mod_up <- gp_mod_up[order(gp_mod_up$p_value), ]
gp_mod_up_BP <- gp_mod_up[gp_mod_up$source == "GO:BP", ]
gp_mod_up_BP$GeneRatio <- paste0(gp_mod_up_BP$intersection_size,  "/", gp_mod_up_BP$query_size)
gp_mod_up_BP$BgRatio <- paste0(gp_mod_up_BP$term_size, "/", gp_mod_up_BP$effective_domain_size)
names(gp_mod_up_BP) <- c("Cluster", "Category", "ID", "Description", "p.adjust", "query_size", "Count", "term_size", "effective_domain_size", "geneID", "GeneRatio", "BgRatio")
gp_mod_up_BP$geneID <- gsub(",", "/", gp_mod_up_BP$geneID)
row.names(gp_mod_up_BP) <- gp_mod_up_BP$ID
gp_mod_enrich_up_BP <- new("enrichResult", result = gp_mod_up_BP)

```
Draw enrichment plot:
```{r warning=FALSE, message=FALSE}
enrichplot::dotplot(gp_mod_enrich_up_BP, showCategory = 10) + ggplot2::labs(title = "up-regulated") + ggplot2::scale_color_gradient(low = "lightseagreen", high = "darkorange1")
```

## Down-regulated proteins

```{r warning=FALSE, message=FALSE}
multi_gp_down_reg <- gost(list("down-regulated" = down_names$name), multi_query = FALSE, evcodes =TRUE)

gp_mod_down = multi_gp_down_reg$result[, c("query", "source", "term_id","term_name", "p_value", "query_size", "intersection_size", "term_size", "effective_domain_size", "intersection")]
gp_mod_down <- gp_mod_down[order(gp_mod_down$p_value), ]

gp_mod_down_BP <- gp_mod_down[gp_mod_down$source == "GO:BP", ]
gp_mod_down_BP$GeneRatio <- paste0(gp_mod_down_BP$intersection_size,  "/", gp_mod_down_BP$query_size)
gp_mod_down_BP$BgRatio <-  paste0(gp_mod_down_BP$term_size, "/", gp_mod_down_BP$effective_domain_size)
names(gp_mod_down_BP) <- c("Cluster", "Category", "ID", "Description", "p.adjust", "query_size", "Count", "term_size", "effective_domain_size", "geneID", "GeneRatio", "BgRatio")
gp_mod_down_BP$geneID <- gsub(",", "/", gp_mod_down_BP$geneID)
gp_mod_enrich_down_BP <- new("enrichResult", result = gp_mod_down_BP)
```

Draw enrichment plot:

```{r warning=FALSE, message=FALSE}
enrichplot::dotplot(gp_mod_enrich_down_BP, showCategory = 10) + ggplot2::labs(title = "down-regulated") + ggplot2::scale_color_gradient(low = "lightseagreen", high = "darkorange1")
```

## biological explanation
Both up- and down-regulated processes exhibit similarities in terms of localization, cellular localization, and the establishment of localization.
"organelle localization", "intracellular protein transport" are down-regulated. They may be disrupted in these cells.
"fatty acid oxidation", "fatty acid beta-oxidation", "fatty acid catabolic processes", and "lipid oxidation"  are up-regulated in CJK cells.The increase in these processes may indicate oxidative stress, which has the potential to stimulate inflammation and calcification pathways.

